Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-03-15 ThesisJP/Background/background.tex
Writing section on adaptive beamforming.

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short impulse consist of many frequency components. A short ultrasound pulse, which would yield high axial resolution, therefore requires a large frequency bandwidth. A ultrasound probe has normally a relative bandwidth between 50 and 100 \%. This corresponds to transmitting a pulse which are between 2 and 1 oscillation long. In the front-end of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{\nom{IQ}{In-phase quadrature (ultrasound data format).} demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. The number of samples in one A-mode image is after this process given by:
\begin{align}
N = \frac{2r}{c}B,
\end{align}
\nomi{$N$}{Number of samples in range}where \nom{$B$}{Frequency bandwidth} is the bandwidth, $B = B_wf_c$ (where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth and \nom{$f_c$}{Center frequency of ultrasound pulse} is the center frequency of the transmitted pulse). The total number of samples in an B-mode image is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
If our cardiac probe has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of element data every $1/55$ second. The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then given as:
\begin{align}
\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}} = MB_wf_c
\end{align}
If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $300$ MFLOPS. Hence, far below the peak processing throughput of modern GPUs and CPUs. The main challange with DAS beamforming is the amount of data that has to be moved. If one element sample is two bytes long, for our example, around 600 MB has to be moved on to the GPU every second. This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even if the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 10.

@@

\subsection{GPUs in medical ultrasound imaging}
Ultrasound imaging systems generates an incredible amount of data. A modern system capable of 3D imaging will generate several tera bytes of data in the front end of the system during a clinical scan. The data rate from a 2D scan is "only" in order giga bytes per second. To handle these amount of data, ultrasound imaging systems have relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline. However, as the computationally power of CPUs and GPUs have increased, ASICs, e.g. for converting data from a polar to a Cartesian grid (known as scan conversion), have gradually been converted into software \cite{Guracar2013}. One of  the last remaining ASICs is the beamformer, whose job is to steer and focus the ultrasound beam in space. This operation involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe. The input data rate is therefor very high. Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are utilizing GPUs in their plane wave beamforming algorithm \cite{Tanter2014}. Beamforming is therefore finally moving into software too. The largest players in the medical ultrasound business, General Electric and Philips, are also well aware of where the trend is going \cite{Thomenius2012} \cite{Metz2011}. Software beamforming opens up new possibilities for advanced processing of ultrasound data. One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}). The increased computational power introduced by GPUs makes it also possible to include advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.

\subsection{Adaptive beamforming}\label{sec:adaptbf}
Adaptive beamforming has several meanings in the medical ultrasound literature. It can for instance refer to aberration correction which is the task of applying time delays to compensate for the distortion introduced by fatty layers in the human body. However, in this thesis adaptive beamforming will refer to data-dependent apodization (or weighting) derived from a given optimization criterion. The optimization criterion which we focus on in this thesis is the minimum variance distortionless response criterion, which is also known as Capon's method \cite{Capon1969}. A beamformer which uses weights derived using Capon's method will in this thesis mainly be referred to as the \textit{Capon beamformer}. However, it is also known as the minimum variance  (\nom{MV}{Minimum variance beamformer}) beamformer or the minimum variance distortionless respons beamformer (\nom{MVDR}{Minimum variance distortionless respons beamformer (Capon)}).

@@

The Capon beamformer produces a weight vector, as introduced in (\ref{eq:das}), based on the impinging signal and an optimization criterion. The problem is as follow \cite{Capon1969}:
\begin{align}
&\min_{\vec{w}} E\{|z[n]|^2\} \rightarrow \min_{\vec{w}} \vec{w}^H \mat{\hat{R}} \vec{w} \label{eq:capon_optimization_criteria} \\
&\text{subjected to } \vec{w}^H\vec{a} = 1,
\end{align}
where \nom{$\mat{\hat{R}}$}{Sample covariance matrix (with regularization)} is a sample covariance matrix. Hence, the resulting weight vector minimizes the output power (the criterion) while maintaining unit gain in the steering direction $\vec{a}$ (distortionless response). The benefit of adaptive beamforming is clearly seen from Fig\,\ref{fig:weights} which shows a sensitivity map (an \textit{array beampattern}) for DAS beamforming with uniform weights and the Capon beamformer. In this simulated case an interfering source is located at 20\degree{} which for the DAS beamformer with uniform weights (\ref{fig:das_weights}) will leak into the beamformer output with its power attenuated with around $13$ dB. Since the Capon weights are data-dependent the method "knows" where the interfering source is located and is able to position a zero in the direction of it. Hence, the interfering source will be highly attenuated independent of its direction. Another interesting case is when the interfering source is located within the mainlobe of the array beampattern. In theory a signal will be regarded by the Capon beamformer as an interfering source as long as the signal propagation vector does not match the steering vector ($\vec{a}$) exactly. If it is not possible to move a zero to the direction of the interferer the mainlobe will get shifted if this results in minimum output power. Adaptive positioning of zeros and mainlobe shifting gives rice to the super-resolution capabilities of the Capon beamformer \cite{Synnevag2007}, which is the main reason why Capon beamforming has been highly studied in the field of medical ultrasound imaging the last decade.

@@

The solution to the minimization problem in (\ref{eq:capon_optimization_criteria}) is, by the method of Lagrange multipliers \cite{VanTrees2003}, found to be
\begin{align}\label{eq:capon_weights}
\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} \in \mathbb{C}^L.
\end{align}
Since $\vec{w}$ is time (or range) dependent one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system. This is indeed computationally demanding, something which becomes clear when we take a closer look at each step of the beamformer. Construction of the covariance matrix has a computationally complexity of $O(L^2N_LN_K)$\nomi{O(.)}{Big O notation (algorithm complexity)}, inverting it has a computationally complexity of $O(L^3)$, and the final normalization and calculation of $\vec{w}$ is $O(L)$. For $L=M/2$, a subarray length often used in the literature, the first two steps both have a complexity of $O(L^3)$. These steps should therefore receive the same amount of attention when a fast implementation of Capon beamforming is attempted. A rough estimate of the number of FLOPS required by a Capon beamformer for real-time cardiac ultrasound imaging is then given by:
\begin{align}
\text{FLOPS}_{\text{Capon}} = L^3N_{\text{B-mode}}f_{\text{FPS}} = L^3B_wf_c.
\end{align}
Which for our example, with $L=M/2=64$, sums up to a required throughput of more than $0.5$ TFLOPS. Even though this is just 10\% of the peak throughput of modern GPUs remember that the latter where only theoretical numbers, and that the actual throughput is highly algorithm dependent. The required throughput for Capon beamforming is, for the moment, hard to achieve in practice \cite{So2011}. As for DAS beamforming this number also scales with higher relative bandwidth, pulse frequency, and more array elements since subarray size typically is proportional to the number of array elements. In the literature on medical ultrasound imaging we find few attempts on implementing a real-time Capon beamformer for medical ultrasound imaging. In addition to the papers in this thesis only Chen \textit{et al.} \cite{Chen2011, Chen, Chen2011a} are known to have explored GPU and FPGA implementations of the Capon beamformer for medical ultrasound imaging, however with some shortcomings  which are discussed in both \textbf{Paper I} and \textbf{II}.

To reduce the processing requirement for Capon beamforming several simplifications have been proposed \cite{Asl2012, Kim}. The most interesting proposals include the low complexity beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace adaptive beamformer by Nilsen and Hafizovic \cite{Nilsen2009}.

Beamspace data is typically refers to the polar grid that cardiac ultrasound data is located in prior to scan conversion. In combination with the Capon beamformer, beamspace refers to the K-space representation of the impinging signals (hence the \nom{FFT}{Fast fourier transform} of the channel data).

@@

The Capon beamformer is derived for narrowband, far-field applications and has typically been applied on medical ultrasound imaging "as is" even tough medical ultrasound imaging is a broadband, near-field imaging application. Later in this section we will explain why a broadband version of the Capon beamformer will have little effect. First we will take a closer look at the details of Capon beamforming.

>>>>

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short impulse consist of many frequency components. A short ultrasound pulse, which would yield high axial resolution, therefore requires a large frequency bandwidth. A ultrasound probe has normally a relative bandwidth between 50 and 100 \%. This corresponds to transmitting a pulse which is between 2 and 1 oscillations when entering tissue. In the front-end of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{\nom{IQ}{In-phase quadrature (ultrasound data format).} demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. The number of samples in one A-mode image is after this process given by:
\begin{align}
N = \frac{2r}{c}B,
\end{align}
\nomi{$N$}{Number of samples in range}where \nom{$B$}{Frequency bandwidth} is the bandwidth, $B = B_wf_c$ (where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth and \nom{$f_c$}{Center frequency of ultrasound pulse} is the center frequency of the transmitted pulse). Note that after this process the ultrasound data will be complex values. The total number of samples in an B-mode image is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
If our cardiac probe has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of element data every $1/55$ second. The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then given as:
\begin{align}
\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}} = MB_wf_c
\end{align}
If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $300$ MFLOPS. Hence, far below the peak processing throughput of modern GPUs and CPUs. The main challange with DAS beamforming is the amount of data that has to be moved. If one element sample is two bytes long, for our example, around 600 MB has to be moved on to the GPU every second. This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 10. It is also exist newer versions of PCI Express which supports even higher transfer rates.

@@

\subsection{GPUs in medical ultrasound imaging}
Ultrasound imaging systems generates an incredible amount of data. A modern system capable of 3D imaging will generate several tera bytes of data in the front end of the system during a clinical scan. The data rate from a 2D scan is "only" in order of giga bytes per second prior to IQ-demodulation. To handle these amount of data, ultrasound imaging systems have relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline. However, as the computationally power of CPUs and GPUs have increased, ASICs, e.g. for converting data from a polar to a Cartesian grid (known as scan conversion), have gradually been converted into software \cite{Guracar2013}. One of  the last remaining ASICs in ultrasound scanners is the beamformer, whose job is to steer and focus the ultrasound beam in space. This operation involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe. The input data rate is therefor very high. Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are for instance utilizing GPUs in their plane wave beamforming algorithm \cite{Tanter2014}. This is the same technique as reported implemented on a GPU by Yiu \textit{et al.} \cite{Yiu2011}. We therfore can conclude that beamforming is finally moving into software too. The largest players in the medical ultrasound business, General Electric and Philips, are also well aware of where the trend is going \cite{Thomenius2012} \cite{Metz2011}. Software beamforming opens up new possibilities for advanced processing of ultrasound data. One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}), and which we will introduce in the next section. The increased computational power introduced by GPUs makes it also possible to include advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.

\subsection{Adaptive beamforming}\label{sec:adaptbf}
Adaptive beamforming has several meanings in the literature on medical ultrasound imaging. It can for instance refer to aberration correction which is the task of applying time delays to compensate for the distortion introduced by fatty layers in the human body. However, in this thesis adaptive beamforming will refer to data-dependent apodization (or weighting) derived from a given optimization criterion. The optimization criterion which we focus on in this thesis is the minimum variance distortionless response criterion, which is also known as Capon's method \cite{Capon1969}. A beamformer which uses weights derived using Capon's method will in this thesis mainly be referred to as the \textit{Capon beamformer}. However, it is also known as the minimum variance  (\nom{MV}{Minimum variance beamformer}) beamformer or the minimum variance distortionless respons beamformer (\nom{MVDR}{Minimum variance distortionless respons beamformer (Capon)}).

@@

The Capon beamformer produces a weight vector, as introduced in (\ref{eq:das}), based on the impinging signal and a constrained optimization problem. The problem is as follow \cite{Capon1969}:
\begin{align}
&\min_{\vec{w}} E\{|z[n]|^2\} \rightarrow \min_{\vec{w}} \vec{w}^H \mat{\hat{R}} \vec{w} \label{eq:capon_optimization_criteria} \\
&\text{subjected to } \vec{w}^H\vec{a} = 1,
\end{align}
where \nom{$\mat{\hat{R}}$}{Sample covariance matrix (with regularization)} is a sample covariance matrix. Hence, the resulting weight vector minimizes the output power (the criterion) while maintaining unit gain in the steering direction $\vec{a}$ (distortionless response). The benefit of Capon beamforming is clearly seen from Fig\,\ref{fig:weights} which shows a sensitivity map (an \textit{array beampattern}) for DAS beamforming with uniform weights and the Capon beamformer. In this simulated case an interfering source is located at 20\degree{} which for the DAS beamformer with uniform weights (\ref{fig:das_weights}) will leak into the beamformer output with its power attenuated with around $13$ dB (first sidelobe). Since the Capon weights are data-dependent the method "knows" where the interfering source is located and is able to position a zero in the direction of it (Fig.\,\ref{fig:capon_weights}). Hence, the interfering source will be highly attenuated independent of its direction. Another interesting case is when the interfering source is located within the mainlobe of the array beampattern. In theory a signal will be regarded by the Capon beamformer as an interfering source as long as the signal propagation vector does not match the steering vector ($\vec{a}$) exactly. If it is not possible to move a zero to the direction of the interferer the mainlobe will get shifted and scaled if this results in minimum output power. Adaptive positioning of zeros and mainlobe shifting gives rice to the super-resolution capabilities of the Capon beamformer \cite{Synnevag2007}, which is the main reason why Capon beamforming has been highly studied in the field of medical ultrasound imaging the last decade.

@@

The solution to the constrained minimization problem in (\ref{eq:capon_optimization_criteria}) is, by the method of Lagrange multipliers \cite{VanTrees2003}, found to be
\begin{align}\label{eq:capon_weights}
\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} \in \mathbb{C}^L.
\end{align}
Since $\vec{w}$ is time (or range) dependent one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system. This is indeed computationally demanding, something which becomes clear when each step of the beamformer is closer examined. It is then found that construction of the covariance matrix has a computationally complexity of $O(L^2N_LN_K)$\nomi{O(.)}{Big O notation (algorithm complexity)}, inverting it has a computationally complexity of $O(L^3)$, and the final normalization and calculation of $\vec{w}$ is $O(L)$. For $L=M/2$, a subarray length often used in the literature, the first two steps both have a complexity of $O(L^3)$. These steps should therefore receive the same amount of attention when a fast implementation of Capon beamforming is attempted. A rough estimate of the number of FLOPS required by a Capon beamformer for real-time cardiac ultrasound imaging is then given by:
\begin{align}
\text{FLOPS}_{\text{Capon}} = L^3N_{\text{B-mode}}f_{\text{FPS}} = L^3B_wf_c.
\end{align}
Which for our example, with $L=M/2=64$, sums up to a required throughput of more than $1/2$ TFLOPS. Even though this is just 10\% of the peak throughput of modern GPUs remember that the latter where only theoretical numbers, and that the actual throughput is highly algorithm dependent. The required throughput for Capon beamforming might, for the moment, be hard to achieve in practice \cite{So2011}. As for DAS beamforming this number also scales with higher relative bandwidth, pulse frequency, and more array elements since the subarray size typically is proportional to the number of array elements. In the literature on medical ultrasound imaging we find few attempts on implementing a real-time Capon beamformer for medical ultrasound imaging. In addition to the papers in this thesis only Chen \textit{et al.} \cite{Chen2011, Chen, Chen2011a} are known to have explored GPU and FPGA implementations of the Capon beamformer for medical ultrasound imaging, however with some shortcomings  which are discussed in both \textbf{Paper I} and \textbf{II}.

To reduce the processing requirement of the Capon beamformer for medical ultrasound imaging several simplified version have been proposed \cite{Asl2012, Kim} in the literature. Among the most interesting proposals we find the low complexity data-dependent beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace (BS) Capon beamformer by Nilsen and Hafizovic \cite{Nilsen2009}. In medical ultrasound imaging beamspace typically refers to the polar grid that cardiac ultrasound data is located in prior to scan conversion. In combination with the Capon beamformer, beamspace refers to the wavenumber-space representation of the recorded signals, hence, the spatial Fourier transform of element data. A GPU implementation of the BS-Capon beamformer for real-time cardiac ultrasound imaging is presented in \textbf{Paper II}. The complexity of BS-Capon beamforming is dominated by the transformation of data into beamspace which has a computational complexity of $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace} is the number of beams selected in beamspace. The initial complexity of BS-Capon beamforming before any tricks are performed is then found to be:
\begin{align}
\text{FLOPS}_{\text{BS-Capon}} = L^2N_bB_wf_c
\end{align}
For our example BS-Capon beamforming, with $L=M/2=64$ and $N_b=3$, will require a throughput of around $30$ GFLOPS. This is a level of processing which both modern CPUs and GPUs should be able to cope with. The rationale behind using the GPU is that a highly parallel task like the Capon beamformer should be offloaded to the processor made for parallel processing. The CPU is then left to do managing and all the task which by tradition occupies it.

@@

The Capon beamformer is derived for narrowband, far-field applications and has typically been applied on medical ultrasound imaging "as is" even tough medical ultrasound imaging is a broadband, near-field imaging application. Later in this section we will explain why a broadband version of the Capon beamformer will have little effect. First we will take a closer look at the details of Capon beamforming.
For the low complexity data-dependent beamformer all the shifted windows are shifted less than the mainlobe width.
